Supplement to ``{\mytitle}''

### 2022-10-10 si/si.Rnw
Updated measurement / rinit models

%%%% parameters %%%%%%%%%%5
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}

@@

We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~34.
% TODO: make sure it aligns with \myeqref{eq:model3:approx}.
The susceptible compartment is initialized so that the sum $S_{u}(0) + I_u(0) + A_u(0) + \sum_k R_{u, k}(0) = \text{population}_u$.
The bacteria compartment is then initialized using Eq.~\myeqref{eq:model3:b0}:
\begin{equation}
  B_u(0) = \big[1 + \seasAmplitude \big(\xi_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0},
\end{equation}
where $\xi_u \in (0, 1)$ is an initial value parameter that allows some flexibility in determining the initial state of the bacteria compartment.

@@

In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}.
\begin{equation}
  \label{model1:meas}
  Y_{n} \mid \Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(n) = z(n) \sim \text{NB}\left(\reportRate z(n), \obsOverdispersion \right),
\end{equation}
where $Y_n$ is the reported cholera cases at time $n \in 1:N$ and $\Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(n)$ is the number of individuals who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $n-1 \in 0:N-1$, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z+ \psi^2\reportRate^2 z^2)$. \jwc{CHECK THIS}.

As mentioned in the main text, Model~2 was fit using log-transformed observed cases.
We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log-transformed cases as the loss function.

\begin{equation}
  \label{model2:meas}
  \log\big(Y_{u, n} + 1\big) \mid \bm{X}^{(2)}(n) = \bm{x}^{(2)}(n) \sim \text{N}\left(\log \big(\reportRate \sum_{\vaccCounter} I_{u, \vaccCounter}(n) + 1\big), \obsOverdispersion^2 \right)
\end{equation}

Where $\sum_{\vaccCounter} I_{u, \vaccCounter}(n)$ is the total number of individuals who are infected at time $n$ in unit $u \in 1:U$, for all vaccination campaigns $\vaccCounter$.
We note that fitting a model with this loss function is equivalent to fitting using least squares, with $\log(y_{u, n} + 1)$ as the response variable.

@@

\subsection{Model~2 Replication}

Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevations charts and river flows.

The source code that was used to implement Model~2 in \cite{lee20} was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}, which facilitates the creation of meta-population models.
We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
instead, we obtained these matrices via personal correspondence with the Model~2 authors.

Using these matrices, and the point estimates for model parameters listed in \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}, suggesting that our re-implementation of Model~2 in the \code{haitipkg} is a faithful approximation to the original model used in \cite{lee20}.

@@

We do note, however, several discrepancies exit between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}, such as a delayed second wave in the epidemic period and a larger initial wave during the endemic period.
While these are only minor discrepancies, because Model~2 is deterministic one may expect to be able to obtain perfect replication in model trajectories.
We believe these discrepancies are a result of slight difference in model parameters.
For example, we note that the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
The collective effect of these small differences in model parameters likely will result in differences in model trajectories, as seen here.

>>>>

%%%% parameters %%%%%%%%%%5
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\Binit{\xi}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}

@@

We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~34.
% TODO: make sure it aligns with \myeqref{eq:model3:approx}.
The susceptible compartment is initialized so that the sum $S_{u}(0) + I_u(0) + A_u(0) + \sum_k R_{u, k}(0) = \text{population}_u$.
The bacteria compartment is then initialized using Eq.~\myeqref{eq:model3:b0}:
\begin{equation}
  B_u(0) = \big[1 + \seasAmplitude \big(\Binit_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0},
\end{equation}
where $\Binit_u \in (0, 1)$ is an initial value parameter that allows some flexibility in determining the initial state of the bacteria compartment.

@@

In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}.
\begin{equation}
  \label{model1:meas}
  Y_{n} \mid \Delta N_{EI}(n) = z(n) \sim \text{NB}\left(\reportRate z(n), \obsOverdispersion \right),
\end{equation}
where $Y_n$ is the reported cholera cases at time $n \in 1:N$ and $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $n-1 \in 0:N-1$, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z+ \psi^2\reportRate^2 z^2)$. \jwc{CHECK THIS}.

\subsection{Model~2}\label{sec:mod2Meas}

As mentioned in the main text, Model~2 was fit using reported case counts that were transformed using the natural logarithm.
We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases as the loss function.
\begin{equation}
  \label{model2:meas}
  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\log \big(\reportRate z(n) + 1\big), \obsOverdispersion^2 \right),
\end{equation}
where $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter$ and unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $n-1 \in 0:N-1$.
Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\mu = \log\big(\Delta N_{EI}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.

This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
\begin{equation}
  \label{model2:meas}
  Y_{u, n} \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\reportRate z(n), \obsOverdispersion^2 \right),
\end{equation}
which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
The measurement model for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
\begin{equation}
  \label{model2:meas}
  Y_{u, n} \mid \Delta N_{EI}(n) = z_1(n), \Delta N_{EA}(n) = z_2(n) \sim \text{N}\left(\reportRate (z_1(n) + z_2), \obsOverdispersion^2 \right),
\end{equation}
where the notation for $\Delta N_{EA}(n)$ is similar to $\Delta N_{EI}(n)$, described above.

@@

\subsection{Model~2 Replication}\label{sec:mod2rep}

Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.

The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}, which facilitates the creation of meta-population models.
We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
instead, we obtained these matrices via personal correspondence with the Model~2 authors.
Using these matrices, and the point estimates for model parameters listed in \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.

@@

There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
% such as a delayed second wave in the epidemic period and a larger initial wave during the endemic period.
While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
In this case, these discrepancies may be attributed to model implementation and small differences in model parameters.
Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
According to repository comments, the files \code{fitInPieces3paramsCleanMay2019Public.py} and \code{fitInPiecesMuWithFracSusFixedAllInfectionsPublic.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results.
These files contain errors that making them impossible to run, and therefore these files cannot be the code that generated the reported model fits\footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{choleraEqsPublic.py}.}.
The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
% The difficulty of reproducing the result of \citet{lee20} is coupled with a misunderstanding of how these results were originally obtained.

\jwc{PROBABLY DELETE THIS NEXT BIT, BUT IT'S USEFUL TO DOCUMENT IT FOR NOW...}
One concern that arises when inspecting the files \code{fitInPieces3paramsCleanMay2019Public.py} and \code{fitInPiecesMuWithFracSusFixedAllInfectionsPublic.py} is that the output of these files (which contains the estimated parameters) was not referenced/used in either the file that generates the sensitivity analysis or the file that generates the model simulations; instead, alternative (non-extant) files with similar names are loaded, suggesting parameters that were used to generate the final results were not a direct result of the estimated parameters.
Furthermore, inspection of the script \code{simulateHaitiFullTimeInPiecesAllVaccinationScenariosSingleParameterVectorAllInfectionsPublic.py}---which is used to obtain simulations for each of the vaccination scenarios---results in a number of concerns.
First, the script uses parameters from the pickle file \code{leastSquaresDataUpTo2019/fitEpidemicPiece\_resultsMatrix\_18Mar2019.pickle}, which only appears in the repository as a commented out line of code in the file \code{generateSetsOfParametersForSensitivityAnalysisPublic.py}, suggesting that the set of parameters used to simulate each vaccination scenario was not the most recent set of parameters generated by the sensitivity analysis.
Second, the script appears to use the set of differential equations from the function \code{choleraEqs11WithVaccinationNetwork} for the epidemic phase, even though parameter values for this phase were estimated using \code{choleraEqs10WithVaccinationNetwork}.
It is possible, however, that we simply are interpreting the code incorrectly and that these concerns are unwarranted.
Furthermore, because the provided code is non-functional, it is possible that the code that was actually used to obtain the reported results does not include these potential errors.
This highlights the importance of providing the exact source code that was used to obtain the reported results of any analysis.

Another potential explanation for the discrepancy between our simulations from Model~2 and those of \citet{lee20} is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}. For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
The collective effect of these small differences in model parameters likely will result in differences in model trajectories.
This issue could have been mitigated if the saved files containing estimated model parameters were included as part of the code submission.
